---
title: "Covid91 vaccine study Final2024"
author: "Matthew Ramirez"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, aids = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
guy = filter(FinalData, sex =="M")
gal = filter(FinalData, sex =="F")
gay = filter(FinalData, LGBTQ =="gay")
Druguser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

# Coding details 

First, the variables from FinalData needed to be changed from character variables into factor variables. the appropriate dplyr code was used to do this. The following code was used to achieve this: 

FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))

Next, specific subgroups had to be separated from the broader categories in order to properly studied. Gal and guy were separated into subgroups, as well as gay sexual orientation and people that answered yes to participating in drug use. All of these categories are isolated to see weather or not these particular populations should  or should not take the  new covid-19 vaccine.


In order to create the effectiveness interval for all four categories, each group of people in this study had to be stored in separate variables. 


## Males

### Barplot



```{r}
dd2 <- guy %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

This plot represents the population of guys(males) people and the relationship to the contraction of covid-19 by either talking the vaccine or the placebo. 


The plot above shows that 25% of the people that took the vaccine did in fact get covid-19, while only approximately 75% of the people taking the placebo contracted covid-19.  A substantially higher percentage of people who didn't take the vaccine contracted covid-19 versus those who were only given the vaccine. The vaccine appears to be effective in protection against covid-19. The odds of getting covid-19 while taking the vaccine is 1:4. The men that took the drug were one and quarter times more likely to get covid-19 than those taking the placebo.


There are two plots. In first plot,the x-axis represents infected, whether the participants were infected with covid-19 or they were not. The y-axis represents the count, or number of people that got covid-19 or not.
In the second plot, the one above, The The left two bars of the graph are all the people that got covid-19. The x-axis represents whether a person got covid-19 or not. The y-axis, however, instead of representing the number of people who got covid-19, now represents the percentage of people that got covid-19 after taking the the placebo or taking the Vaccine. 


The codes that  create this plot start with the code block beginning with dd2. dd2 can be seen in the environment pane and if inspected contains 4 variables.These variables include infected, treatment, count, and prcent(stands for percent). The second group of codes is called the basicC. BasicC can also be seen in the environment pane and appears to contain a list of 9 variables. This information creates the first plot. The basicCC is the last bunch of codes and is also in the files pane. It also contains 9 variables and helps to create the percentage plot, the second plot.

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=guy)
rowPerc(table1)
colPerc(table1)
```

These numerical results show specific percentages of guys(men) that got covid-19 after taking the placebo and taking the vaccine. 
Among the guys, 74% got covid-19 after taking the vaccine, while the approximately 26% of guys that were given the placebo contracted covid-19. The vaccine is clearly not effective in preventing covid-19 in guys(males) and it is not recommended. Those who took the placebo were much less likely to contact covid-19
The odds of getting covid-19 while taking the vaccine is 1:4. The men that took the drug were one and quarter times more likely to get covid-19 than those taking the placebo.


### Inferential 



```{r}
chisq.test(table1)
chisqtestGC(table1)
guysfish <- fisher.test(table1)
guysfish
```


The information to play close attention to among all of these results is the output of the Fisher's Exact test. The p-value displayed above is 2.2e-16. This is a small number. Seeing this number means that these results are not likely to occur by chance.



The null hypothesis predicted that the deaths related to covid-19 are independent of sex. Since it is clear that sex does play a role in the death rate of individuals, the alternative hypothesis, that there is a correlation between covid-19 deaths and sex, is the favorable option.

The odd ratio above illustrates a ratio of 2.815. the guys(men) who received that vaccine were 3 times more likely to get covid-19 fromd the placebo. 


```{r}
effect <- guysfish$p.value/(guysfish$p.value +1)
effectivenessguy = guysfish$estimate/(guysfish$estimate +1)*100
print(effectivenessguy)
```

```{r}
low=(guysfish$conf.int[1]/(guysfish$conf.int[1] +1))*100
hi=(guysfish$conf.int[2]/(guysfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`



Above is the effectiveness for the covid-19 vaccine for guys(males). If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we cn conclude that the vaccine is effective in protecting guys from getting covid-19. 

The odds ratio is approximately how effective the vaccine is. The odds ratio for the guys can be seen above as 73. The effectiveness confidence interval for the guys is 2.2 to 3.6.

## Females

### Barplot

```{r}
dd2 <- gal %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```


This plot represents the population of gals(females) people and the relationship to the contraction of covid-19 by either talking the vaccine or the placebo. 

In this plot, contrary to the previous, it appears that more people who were only given the placebo contracted covid-19. Only approximately 35% of the individuals who were given the vaccine were reported to have contracted covid-19. That means around 65% of the people that got covid-19 were the individuals that only received the placebo.The odds of getting covid-19 while taking the vaccine is 1:3. The gals that took the vaccine were half as likely to get covid-19 than those taking the placebo.


There are two plots. In first plot,the x-axis represents infected, whether the participants were infected with covid-19 or they were not. The y-axis represents the count, or number of people that got covid-19 or not.
In the second plot, the one above, The The left two bars of the graph are all the people that got covid-19. The x-axis represents whether a person got covid-19 or not. The y-axis, however, instead of representing the number of people who got covid-19, now represents the percentage of people that got covid-19 after taking the the placebo or taking the Vaccine. 



The codes that  create this plot start with the code block beginning with dd2. dd2 can be seen in the environment pane and if inspected contains 4 variables.These variables include infected, treatment, count, and prcent(stands for percent). The second group of codes is called the basicC. BasicC can also be seen in the environment pane and appears to contain a list of 9 variables. This information creates the first plot. The basicCC is the last bunch of codes and is also in the files pane. It also contains 9 variables and helps to create the percentage plot, the second plot.


### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=gal)
rowPerc(table1)
colPerc(table1)
```

These numerical results show specific percentages of gals(women) that got covid-19 after taking the placebo and taking the vaccine. 
The gals(women) that took the vaccine ans contracted covid-19 was 35%, just as approximated in the plots prior. The gals that took the placebo and got covid-19 was 65%. The vaccine is in fact effective in protecting gals against the contraction of covid-19. 
The odds of getting covid-19 while taking the vaccine is 1:2 for gals. The gals that took the vaccine were half as likely to get covid-19 than those taking the placebo.



### Inferential 




```{r}
chisq.test(table1)
chisqtestGC(table1)
galsfish <- fisher.test(table1)
galsfish
```



The information to play close attention to among all of these results is the output of the Fisher's Exact test. The p-value displayed above is extremely small at 1e-08. This is an extremely tiny number. seeing this number means that these results are not likely to occur by chance and so the null hypothesis can be rejected in favor of the alternative.


The Fishers's Exact Test illustrates that the odds ratio is 1.9.
gals(Women) who got the drug were about 0.49 times more likely to get covid-19 than women that only took the placebo.


```{r}
effect <- galsfish$p.value/(galsfish$p.value +1)
effectivenessgal = galsfish$estimate/(galsfish$estimate +1)*100
print(effectivenessgal)
```

```{r}
low=(galsfish$conf.int[1]/(galsfish$conf.int[1] +1))*100
hi=(galsfish$conf.int[2]/(galsfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`

Above is the effectiveness for the covid-19 for gals(females). If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we cn conclude that the vaccine is effective in protecting gals from getting covid-19. 

The hallmark of an effective vaccine is 65% or better. This vaccine works great in protecting gals(women) from contructing covid-19.

The odds ratio is approximately how effective the vaccine is. The odds ratio for the gals can be seen above as 65. The effectiveness confidence interval for the gals is 1.5 to 2.3.

## LGBTQ

### Barplot

```{r}
dd2 <- gay %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```



This plot represents the population of gay people and the relationship to the contraction of covid-19 by taking either the vaccine or the placebo.

In this barplot, it appears that a sniffling 20% of the individuals that took the vaccine got covid-19 as well as the 80% who received the placebo. The odds of getting covid-19 while taking the vaccine is 1:5. The individuals who identified as gay and took the vaccine were 5 times as likely to not get covid-19 than those taking the placebo.


There are two plots. In first plot,the x-axis represents infected, whether the participants were infected with covid-19 or they were not. The y-axis represents the count, or number of people that got covid-19 or not.
In the second plot, the one above, The The left two bars of the graph are all the people that got covid-19. The x-axis represents whether a person got covid-19 or not. The y-axis, however, instead of representing the number of people who got covid-19, now represents the percentage of people that got covid-19 after taking the the placebo or taking the Vaccine. 


The codes that  create this plot start with the code block beginning with dd2. dd2 can be seen in the environment pane and if inspected contains 4 variables.These variables include infected, treatment, count, and prcent(stands for percent). The second group of codes is called the basicC. BasicC can also be seen in the environment pane and appears to contain a list of 9 variables. This information creates the first plot. The basicCC is the last bunch of codes and is also in the files pane. It also contains 9 variables and helps to create the percentage plot, the second plot.

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data=gay)
rowPerc(table1)
colPerc(table1)
```

These numerical results show specific percentages of gay individuals that got covid-19 after taking the placebo and taking the vaccine. 
The gay participants that took the vaccine and contracted covid-19 was at an approximately 20%! Those participants that only received the placebo only contacted covid-19 at 80%. The vaccine id evidently completely effective in protecting the gay community against covid-19 and taking it is advised. 
The odds of getting covid-19 while taking the vaccine is 1:5. The individuals who identified as gay and took the vaccine were .05 times as likely to get covid-19 than those taking the placebo.





### Inferential 

```{r}
chisq.test(table1)
chisqtestGC(table1)
gaysfish <- fisher.test(table1)
gaysfish
```


The p-value is 7e-07. This in an incredibly small number and with that being said, the conclusion is that these results are unlikely to occur by chance.

The Fishers's Exact Test illustrates that the odds ratio is 4.3.
Gay individuals who got the vaccine were about .05 times more likely to get Covid-19 than gays that took the placebo.

The null hypothesis is rejected in favor the the alternative. The drug is completely ineffective in preventing the chances of getting covid-19 in individuals who identify as gay.





```{r}
effect <- gaysfish$p.value/(gaysfish$p.value +1)
effectivenessgays = gaysfish$estimate/(gaysfish$estimate +1)*100
print(effectivenessgays)
```

```{r}
low=(gaysfish$conf.int[1]/(gaysfish$conf.int[1] +1))*100
hi=(gaysfish$conf.int[2]/(gaysfish$conf.int[2] +1))*100
options(digits=2)
```





## Effectiveness Confidence Interval `r low`  to `r hi`

Above is the effectiveness for the covid-19 for those who identified as gay.  If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we can conclude that the vaccine is ineffective in protecting people who identify as gay from getting covid-19. 

The odds ratio is approximately how effective the vaccine is. The odds ratio for the gays folk be seen above as 80. This is completely effective. The people who take the vaccine are better off then those who don't receive it at all. The effectiveness confidence interval for the people who identified as gay is 2.2 to 8.9.

## Druggies

### Barplot


```{r}
dd2 <-  Druguser %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

This plot represents the population of people who answered  Druguser to drug use and the relationship to the contraction of covid-19 while taking either the vaccine or the placebo.

This barplot shows that arounf 89% of the people who answered yes to drug use and recieved the vaccine contracted covid-19. This means that around 11% of the drugusers that only took the placebo ended up getting covid-19.The odds of getting covid-19 while taking the vaccine is 8:1. The drugusers that took the vaccine were very likely to get covid-19 than those taking the placebo.




There are two plots. In first plot,the x-axis represents infected, whether the participants were infected with covid-19 or they were not. The y-axis represents the count, or number of people that got covid-19 or not.
In the second plot, the one above, The The left two bars of the graph are all the people that got covid-19. The x-axis represents whether a person got covid-19 or not. The y-axis, however, instead of representing the number of people who got covid-19, now represents the percentage of people that got covid-19 after taking the the placebo or taking the Vaccine. 




The codes that  create this plot start with the code block beginning with dd2. dd2 can be seen in the environment pane and if inspected contains 4 variables.These variables include infected, treatment, count, and prcent(stands for percent). The second group of codes is called the basicC. BasicC can also be seen in the environment pane and appears to contain a list of 9 variables. This information creates the first plot. The basicCC is the last bunch of codes and is also in the files pane. It also contains 9 variables and helps to create the percentage plot, the second plot.

### Numerical results

```{r}
table1 <- xtabs(~infected + treatment, data= Druguser)
rowPerc(table1)
colPerc(table1)
```
 

These numerical results show specific percentages of drugusers that got covid-19 after taking the placebo and taking the vaccine.
The percentage of drug users that took the vaccine and contacted covid-19 is 92% Those that were only given the placebo and got covid-19 was almost 8%. It is clear that the vaccine to prevent covid-19 is not very effective in protecting drug users against covid-19 and so getting this vaccine is not recommended. The odds of getting covid-19 while taking the vaccine is 9:1. The drugusers that took the vaccine were as very likely to get covid-19 than those taking the placebo.


### Inferential 


```{r}
chisq.test(table1)
chisqtestGC(table1)
drugfish <- fisher.test(table1)
drugfish
```


The p-value is 3e-12. This in a small number and with that being said, the conclusion is that these results are unlikely to occur by chance.

The Fishers's Exact Test illustrates that the odds ratio is 0.087
drug users who got the vaccine were about half as likely to get Covid-19 than drug users that took the placebo.

The null hypothesis is rejected in favor the the alternative. The drug is not effective in preventing the chances of getting covid-19 in drugusers.




```{r}
effect <- drugfish$p.value/(drugfish$p.value +1)
effectivenessdrug = drugfish$estimate/(drugfish$estimate +1)*100
print(effectivenessdrug)
```

```{r}
low=(drugfish$conf.int[1]/(drugfish$conf.int[1] +1))*100
hi=(drugfish$conf.int[2]/(drugfish$conf.int[2] +1))*100
options(digits=2)
```

## Effectiveness Confidence Interval `r low`  to `r hi`


Above, is the effectiveness for the covid-19 for guys. If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we can conclude that the vaccine is not very effective in protecting drug users from getting covid-19. 

The hallmark of an effective vaccine is 8%. This vaccine does not work greatly in protecting drug users from contracting covid-19.

The odds ratio is approximately how effective the vaccine is. The odds ratio for the drug users can be seen above as 8. The effectiveness confidence interval for the drug users is 0.027 to 0.217.

# Overall Results and Conclusions

The covid-19 vaccine was effective in protecting men against the contraction of the disease. The covid-19 effectiveness was `r effectivenessguy`
for men.
The covid-19 vaccine was pretty successful in protecting women against covid-19.The covid-19 effectiveness was `r effectivenessgal`. women were protected the third most with this particular vaccine.
The covid-19 vaccine was very effective in treating people who identified as gay. The individuals who identified as gay and took the vaccine were 0.05 times as likely to get covid-19 than those taking the placebo.The covid-19 effectiveness was `r effectivenessgays` for gay individuals.
The individuals that were drugusers were not effectively protected by the vaccine though not quite as much as the gals(women), guys(men), and Gays(LGBTQ). 
The drugusers that took the vaccine were the most likely to get covid-19 than those taking the placebo.The covid-19 effectiveness was `r effectivenessdrug` for drugusers.

If you are a male or identify as gay it is recommended to receive the vaccine. However, if you are a female or a drug user, the vaccine was reported to have been not been as effective or almost not effective.
The null that there is no difference in taking the vaccine can be accepted to the categories including drug users. the alternative, that the vaccine works and is effective, can be accepted for the gals(women), guys(men), and gays(LGBTQ).



